Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-01-24 Manuscript/Figures/figure_legends.Rmd
edits

**Figure 5. Predicted TAD boundaries are enriched with known molecular drivers of the 3D chromatin architecture.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17390000-18640000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox. Called TADs are outlined in blue. The blue vertical lines represent a called TAD boundary identified by the ARROWHEAD algorithm. The black line represents the predicted probability of each base pair being a TAD boundary. Regions of adjacent base pairs with probability of 1 are represented by the green regions along the x-axis. The green vertical lines represent the subsequent base pair predicted to be a TAD boundary using medoid partitioning. <!--(B) Enrichment heatmaps illustrate the average signal strength of ChIP-seq peak regions defined by SMC3, ZNF143, CTCF, and RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.-->

>>>>

**Figure 5. Predicted TAD boundaries are enriched with known molecular drivers of the 3D chromatin architecture.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17389000-18011000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox. Called TADs are outlined in blue. The blue vertical lines represent a called TAD boundary identified by the ARROWHEAD algorithm. The black line represents the predicted probability of each base pair being a TAD boundary. Regions of adjacent base pairs with probability of 1 are represented by the green regions along the x-axis. The green vertical lines represent the subsequent base pair predicted to be a TAD boundary using medoid partitioning. <!--(B) Enrichment heatmaps illustrate the average signal strength of ChIP-seq peak regions defined by SMC3, ZNF143, CTCF, and RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.-->

### 2020-01-24 Manuscript/manuscript.Rmd
edits

Given that many TFBS are known to colocalize together on the linear genome and are therefore highly correlated, we sought to further optimize our TAD prediction model to avoid overfitting by performing recursive feature elimination. We found that, under our framework, we were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figure 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine the most common TFBS, we clustered the predictive importances across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions for all chromosomes (Figure 4C; Supplementary Figure 4C). Therefore, using our framework *TADRF* was able to identify the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self].

@@

After obtaining our optimally built *TADRF* model, we investigated whether we could leverage our model to alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. Our method, referred to as *preciseTAD*, uses a combination of complete linkage hierarchical clustering with a scalable version of partitioning around mediods (PAM), known as clustering large applications (CLARA) to reduce many adjacent base pairs with similarly high predictive probabilities down to a set of the most representative and biologically meaningful predicted TAD boundaries (see Methods). *preciseTAD* was applied to chromosome-specific sections of the linear genome. For each chromosome, we defined our new validation data using distance-type predictors among only the transcription factor binding sites for each base pair. That is, each base pair had a corresponding row vector, **X**=$\{X_{1}, ..., X_{p}\}$, representing the $log_{2}$ genomic distances to the nearest respective transcription factor binding sites (Supplementary Table 2).

@@

<!--
![**Predicted TAD boundaries are enriched in known TAD boundary-forming TFBSs.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17390000-18640000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox.  Called TADs are outlined in blue.  The blue vertical lines represent a called TAD boundary identified by the ARROWHEAD algorithm. The black line represents the predicted probability of each base pair being a TAD boundary. Regions of adjacent base pairs with probability of 1 are represented by the green regions along the x-axis. The green vertical lines represent the subsequent base pair predicted to be a TAD boundary using medoid partitioning. (B) Enrichment heatmaps illustrate the average signal strength of ChIP-seq peak regions defined by SMC3, ZNF143, CTCF, and RAD21 around the centers of 10 kb flanked predicted and called TAD boundaries.](Figures/visualizing_boundaries_example_with_enriched_heatmaps.png) 
-->

>>>>

Given that many TFBS are known to colocalize together on the linear genome and are therefore highly correlated, we sought to further optimize our TAD prediction model to avoid overfitting by performing recursive feature elimination. We found that, under our framework, we were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figure 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine the most common TFBS, we clustered the predictive importances across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions for all chromosomes (Figure 4C; Supplementary Figure 4C). Therefore, using our framework *TADRF* was able to identify the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self]. Therefore, we could further optimize our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. To summarize, the optimally built cell line-specific random forest model under our *TADRF* framework was found to be one built using 10 kb resolution data, leveraging distance-type predictors for the CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites, while implementing random under-sampling to balance TAD and non-TAD boundary defined genomic bins.

@@

After obtaining our optimally built *TADRF* model, we investigated whether we could leverage our model to alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. Our method, referred to as *preciseTAD*, uses a combination of complete linkage hierarchical clustering with a scalable version of partitioning around mediods (PAM), known as clustering large applications (CLARA) to reduce many adjacent base pairs with similarly high predictive probabilities down to a set of the most representative and biologically meaningful predicted TAD boundaries (see Methods). *preciseTAD* was applied to chromosome-specific sections of the linear genome. For each chromosome, we defined our new validation data using distance-type predictors among only the most informative TFBS for each base pair, namely the CTCF, RAD21, SMC3, and ZNF143 transcription factors. That is, each base pair had a corresponding row vector, **X**=$\{X_{1}, ..., X_{4}\}$, representing the $log_{2}$ genomic distances to the nearest respective transcription factor binding site.

@@

![**Predicted TAD boundaries are enriched in known TAD boundary-forming TFBSs.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17389000-18011000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox.  Called TADs are outlined in blue.  The blue vertical lines represent a called TAD boundary identified by the ARROWHEAD algorithm. The black line represents the predicted probability of each base pair being a TAD boundary. Regions of adjacent base pairs with probability of 1 are represented by the green regions along the x-axis. The green vertical lines represent the subsequent base pair predicted to be a TAD boundary using medoid partitioning.](Figures/visualizing_ptbp_and_ctbp.png)
